Well-quasi-orderings on word languages

### 2025-10-15 src/appendix.tex
new: moving stuff to appendix (#10)
Co-authored-by: Aliaume LOPEZ <lopezaliaume@gmail.com>
hotfix: restoring stuff

\section{Proofs for Section~\ref{prefixes:sec}}
\begin{figure}
    \centering
    \includestandalone{fig/antichain-branch-standalone}
    \caption{An \kl{antichain branch} for the language $a^* b$,
        represented in the \kl{tree of prefixes} over the alphabet $\set{a,b}$.
        The branch is represented with dashed lines in turquoise, and the
        \kl{antichain} is represented in dotted lines in blood-red.
    }
    \label{antichain-branch:fig}
\end{figure}

@@

@@

>>>>

\section{Proofs for Section~\ref{prefixes:sec}}

\begin{figure}
    \centering
    \includestandalone{fig/antichain-branch-standalone}
    \caption{An \kl{antichain branch} for the language $a^* b$,
        represented in the \kl{tree of prefixes} over the alphabet $\set{a,b}$.
        The branch is represented with dashed lines in turquoise, and the
        \kl{antichain} is represented in dotted lines in blood-red.
    }
    \label{antichain-branch:fig}
\end{figure}

@@

\begin{lemma}
	\label{automatic-uur:lem}
	\proofref{automatic-uur:lem}
	Given an \kl{automatic sequence} $w \in \Sigma^{\Nat}$, one can decide
	whether it is \kl{ultimately uniformly recurrent}.
\end{lemma}

\begin{proofof}{automatic-uur:lem}
	We can rewrite this as a question on the \kl{automatic sequence} $w$
	as follows:
	\begin{align*}
		&\exists N_0,                   &   \text{ultimately} \\
		&\forall i_s \geq N_0,          &   \text{for every infix (start) } u \\
		&\forall i_e > i_s,             &   \text{for every infix (end) }   u \\
		&\exists k \geq 1,              &   \text{there exists a bound} \\
		&\forall j_s \geq N_0,          &   \text{for every other infix (start) } v \\
		&\forall j_e \geq j_s + k,      &   \text{of size at least $k$} \\
		&\exists l \geq 0,              &   \text{there exists a position in } v \\
		&\forall 0 \leq m < i_e - i_s,  &   \text{where } u \text{ can be found} \\
		&j_s + m + l < j_e \land
		w(i_s + m) = w(j_s + m + l) \quad .
	\end{align*}
	Because $w$ is computable by a finite automaton, one can reduce the above
	formula to a regular language, for which it suffices to check emptiness, which
	is decidable.
\end{proofof}

\begin{proofof}{small-ordinal-invariants:lem}
	Let $N_0$ be a bound such that $w_{\geq N_0}$ is \kl{uniformly recurrent}.
	Let us write $\infset{w}$ the set of finite infixes of $w$.
	We prove that $\oWidth{\infset{w}} \leq \omegaOrd + N_0$.
	Let $u_1 \infleq w$ be a finite word.

	If $u_1$ is an \kl{infix} of $w_{\geq N_0}$, then there exists $k \geq 1$
	such that $u_1$ is an \kl{infix} of every word of size at least $k$. In
	particular, there is finite bound on the length of every sequence of
	incomparable elements starting with $u_1$. We conclude in particular that
	$\infset{w} \setminus \upset{u_1}$ has a finite \kl{ordinal width}.

	Otherwise, $u_1$ can only be found \emph{before} $N_0$. In this case, we
	consider a second element of a \kl{bad sequence} $u_2 \infleq w$, which is
	incomparable with $u_1$ for the \kl{infix relation}. If $u_2$ is an
	\kl{infix} of $w_{\geq N_0}$, then we can conclude as before. Otherwise,
	notice that $u_1$ and $u_2$ cannot start at the same position in $w$
	(because they are incomparable). Continuing this argument, we conclude that
	there are at most $N_0$ elements starting before $N_0$
	at the start of any sequence of
	incomparable elements in $\infset{w}$. We conclude that
	$\oWidth{\infset{w}} \leq \omegaOrd + N_0$.

	Let us now justify that this bound is tight.
	The \kl{Thue-Morse sequence} over a binary
	alphabet $\set{a,b}$ has \kl{ordinal width} $\omegaOrd$
	from \cref{thue-morse-ordinal:lemma}.
	Given a number $N_0
	\in \Nat$, one can construct an arbitrarily long \kl{antichain} of words for
	the \kl{infix relation} by using a new letter $c$. When concatenating this
	(finite) antichain as a prefix of the \kl{Thue-Morse sequence}, one obtains a
	new (infinite) word $w$. It is clear that the \kl{ordinal width} of
	$\infset{w}$ is now at least $\omegaOrd + N_0$.
\end{proofof}

@@

For this paper to be self-contained, we will also recall how runs of a finite
state automaton can be understood as an \kl{amalgamation system}.

\begin{example}[{\cite[Section 3.2]{ASZZ24}}]
  \label{automaton-amalgamation:example}
    Let $A = (Q, \delta, q_0, F)$ be a finite state automaton over a finite
    alphabet $\Sigma$. Let $\Delta$ be the set of transitions $(q_1, a, q_2)
    \in Q \times \Sigma \times Q$,
    and $R \subseteq \Delta^*$ be the set of 
    words over transitions that start with the initial state $q_0$,
    end in a final state $q_f \in F$, and such that the end state of a
    letter is the start state of the following one.
    The canonical decomposition $\canrun$
    is defined as a morphism from $\Delta^*$ to $\Sigma^*$
    that maps $(q,a,p)$ to $a$. 
    Because of the one-to-one correspondence of steps of a run $\rho$ and letters in its \kl{canonical decomposition}, 
    we may treat the two interchangeably.
    Finally, given two runs $\rho$ and $\sigma$ of the automaton,
    we say that an embedding $f \in \HigEmb(\canrun(\rho), \canrun(\sigma))$
    belongs to $E(\rho,\sigma)$ when
    $f$ is also defining an embedding from $\rho$ to $\sigma$ as words in $\Delta^*$.

    The system $(\Sigma, R, E, \canrun)$ is an \kl{amalgamation system},
    whose language is precisely the language of words recognized
    by the automaton $A$.
\end{example}
\begin{proof}
    By definition, the embeddings inside $E(\rho,\sigma)$ are in
    of $\HigEmb(\canrun(\rho), \canrun(\sigma))$, and they compose properly.
    Because $\Delta = Q \times \Sigma \times Q$ is finite, it is 
    a \kl{well-quasi-ordering} when equipped with the equality relation, and 
    we conclude that $\Delta^*$ with $\higleq$ is a \kl{well-quasi-order}
    according to Higman’s Lemma \cite{HIG52}.

    Let us now move to proving that the system satisfies the amalgamation
    property. Given three runs $\rho,\sigma,\tau \in R$, and two embeddings $f \in E(\rho,\sigma)$
    and $g \in E(\rho,\tau)$, we want to construct an amalgamated run $\sigma \vee \tau$.
    Because letters in the run $\rho$ respect the transitions of the automaton
    (i.e., if the letter $i$ ends in state $q$, then the letter $i+1$ starts in
    state $q$), then the \kl{gap word} at position $i$ starts in state $q$ and
    ends in state $q$ too. This means that for both embeddings
    $f$ and $g$, the \kl{gap words} are read by the automaton by looping
    on a state. In particular, these loops can be taken in any order
    and continue to represent a valid run. That is, we can even select
    the order of concatenation in the amalgamation for \emph{all} 
    $0 \leq i \leq \card{\canrun(\rho)}$ and not just for one separately.

    We conclude by remarking that 
    the language of this amalgamation system is
    the set of $\yieldrun(R)$, 
    because $R$ is the set of valid runs of the automaton,
    and $\yieldrun(\rho)$ is the word read along a run $\rho$.
\end{proof}
